---
title: "Univariate and multivariable regression"
subtitle: EPIR Handbook - Chapter 19
author: "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '4'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scrool: no
    theme: readable
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(
  rio,          # File import
  here,         # File locator
  tidyverse,    # data management + ggplot2 graphics, 
  stringr,      # manipulate text strings 
  purrr,        # loop over objects in a tidy way
  gtsummary,    # summary statistics and tests 
  broom,        # tidy up results from regressions
  lmtest,       # likelihood-ratio tests
  parameters,   # alternative to tidy up results from regressions
  see,          # alternative to visualise forest plots
  flextable     # converting tables to pretty images
  )
```

# Intro

1. Univariate: two-by-two tables
2. Stratified: mantel-haenszel estimates
3. Multivariable: variable selection, model selection, final table
4. Forest plots

```{r import-data}
# import the linelist
linelist <- import("data/linelist_cleaned.rds")
```

# Clean data

```{r define-interest-variables}
## define variables of interest 
explanatory_vars <- c("gender", "fever", "chills", "cough", "aches", "vomit")

## convert dichotomous variables to 0/1 
linelist <- linelist %>%  
  mutate(across(                                      
    .cols = all_of(c(explanatory_vars, "outcome")),  ## for each column listed and "outcome"
    .fns = ~case_when(                              
      . %in% c("m", "yes", "Death")   ~ 1,           ## recode male, yes and death to 1
      . %in% c("f", "no",  "Recover") ~ 0,           ## female, no and recover to 0
      TRUE                            ~ NA_real_)    ## otherwise set to missing
    )
  )

## add in age_category to the explanatory vars 
explanatory_vars <- c(explanatory_vars, "age_cat")

## drop rows with missing information for variables of interest 
linelist <- linelist %>% 
  drop_na(any_of(c("outcome", explanatory_vars)))
```

# Univariate

Two options for doing univariate analysis:

- Use functions available in **base** R to quickly print results to the console. Use the `broom` package to tidy up the outputs.
- Use the `gtsummary` package to model and get publication-ready outputs

## **base** R

### Linear regression {.tabset}

The base R function lm() perform linear regression, assessing the relationship between numeric response and explanatory variables that are assumed to have a linear relationship.

```{r linear-regression}
lm_results <- lm(ht_cm ~ age, data = linelist)
```

#### Summary

run summary() on the model results to see the coefficients (Estimates), P-value, residuals, and other measures.

```{r summary-lm}
summary(lm_results)
```

#### tidy
use the tidy() function from the broom package to pull the results in to a table. for each year increase in age the height increases by 3.5 cm and this is statistically significant

```{r}
tidy(lm_results)
```

#### Visualization

```{r visulize-lm}
## pull the regression points and observed data into one dataset
points <- augment(lm_results)

## plot the data using age as the x-axis 
ggplot(points, aes(x = age)) + 
  ## add points for height 
  geom_point(aes(y = ht_cm)) +
  ## add your regression line 
  geom_line(aes(y = .fitted), colour = "red")
```

### [Logistic regression](https://epirhandbook.com/univariate-and-multivariable-regression.html#logistic-regression) 

The function `glm()` from the stats package (part of base R) is used to fit *Generalized Linear Models (GLM)*.

- `formula = ` The model is provided to `glm()` as an equation, with the outcome on the left and explanatory variables on the right of a tilde `~`.
- `family = ` This determines the type of model to run. For logistic regression, use `family = "binomial"`, for poisson use `family = "poisson"`. Other examples are in the table below.
- `data = ` Specify your data frame

#### Univariate `glm()`

```{r univariate-glm}
model <- glm(outcome ~ age_cat, family = "binomial", data = linelist) %>% 
  tidy(exponentiate = TRUE, conf.int = TRUE) %>%        # exponentiate and produce CIs
  mutate(across(where(is.numeric), round, digits = 2))  # round all numeric columns

counts_table <- linelist %>% 
  janitor::tabyl(age_cat, outcome)

combined <- counts_table %>%           # begin with table of counts
  bind_cols(., model) %>%              # combine with the outputs of the regression 
  select(term, 2:3, estimate,          # select and re-order cols
         conf.low, conf.high, p.value) %>% 
  mutate(across(where(is.numeric), round, digits = 2)) ## round to 2 decimal places

combined %>% 
  flextable::qflextable()
```

## **gtsummary** package

```{r gtsummary-glm}
# theme_gtsummary_journal(journal = "jama")

univ_tab <- linelist %>% 
  dplyr::select(explanatory_vars, outcome) %>% ## select variables of interest

  tbl_uvregression(                         ## produce univariate table
    method = glm,                           ## define regression want to run (generalised linear model)
    y = outcome,                            ## define outcome variable
    method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
    exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
  )

## view univariate results table 
univ_tab

# reset gtsummary themes
# reset_gtsummary_theme()
```



# Multivariable

## Conductmultivriable
add more variables to the right side of the equation, separated by plus symbols (+).

```{r}
mv_reg <- glm(outcome ~ gender + fever + chills + cough + aches + vomit + age_cat, family = "binomial", data = linelist)

summary(mv_reg)
```

If you want to include two variables and an interaction between them you can separate them with an asterisk * instead of a +. Separate them with a colon : if you are only specifying the interaction. 
```{r}
glm(outcome ~ gender + age_cat * fever, family = "binomial", data = linelist)
```

## Combine with **gtsummary**
The gtsummary package provides the tbl_regression() function, which will take the outputs from a regression (glm() in this case) .

```{r}
## run a regression with all variables of interest 
mv_reg <- explanatory_vars %>%  ## begin with vector of explanatory column names
  str_c(collapse = "+") %>%     ## combine all names of the variables of interest separated by a plus
  str_c("outcome ~ ", .) %>%    ## combine the names of variables of interest with outcome in formula style
  glm(family = "binomial",      ## define type of glm as logistic,
      data = linelist)          ## define your dataset

## choose a model using forward selection based on AIC
## you can also do "backward" or "both" by adjusting the direction
final_mv_reg <- mv_reg %>%
  step(direction = "forward", trace = FALSE)

## show results table of final regression 
mv_tab <- tbl_regression(final_mv_reg, exponentiate = TRUE)

mv_tab
```

```{r}
## combine with univariate results 
tbl_merge(
  tbls = list(univ_tab, mv_tab),                          # combine
  tab_spanner = c("**Univariate**", "**Multivariable**")) # set header names
```

# Forest plot

## **ggplot2** package

You can build a forest plot with ggplot() by plotting elements of the multivariable regression results. Add the layers of the plots using these “geoms”:

- estimates with geom_point()
- confidence intervals with `geom_errorbar()`
- a vertical line at OR = 1 with `geom_vline()`

```{r}
options(scipen=999)

mv_tab_base <- final_mv_reg %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%  ## get a tidy dataframe of estimates 
  mutate(across(where(is.numeric), round, digits = 2))          ## round 
```

```{r}
## remove the intercept term from your multivariable results
mv_tab_base %>% 
  
  #set order of levels to appear along y-axis
  mutate(term = fct_relevel(
    term,
    "vomit", "gender", "fever", "cough", "chills", "aches",
    "age_cat5-9", "age_cat10-14", "age_cat15-19", "age_cat20-29",
    "age_cat30-49", "age_cat50-69", "age_cat70+")) %>%
  
  # remove "intercept" row from plot
  filter(term != "(Intercept)") %>% 
  
  ## plot with variable on the y axis and estimate (OR) on the x axis
  ggplot(aes(x = estimate, y = term)) +
  
  ## show the estimate as a point
  geom_point() + 
  
  ## add in an error bar for the confidence intervals
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high)) + 
  
  ## show where OR = 1 is for reference as a dashed line
  geom_vline(xintercept = 1, linetype = "dashed")
```
## **easystats** package

```{r}
pacman::p_load(easystats)
```


