---
title: "Descriptive tables"
subtitle: "EPIR Handbook - Chapter 17"
author: 
  - "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scrool: false
    theme: paper
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
pacman::p_load(
  rio,          # File import
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  gtsummary,    # summary statistics and tests
  rstatix,      # summary statistics and statistical tests
  janitor,      # adding totals and percents to tables
  scales,       # easily convert proportions to percents  
  flextable     # converting tables to pretty images
  )

knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(htmltools.dir.version = FALSE)
```

Consider the points below as you choose the tool for your situation.

- Use `tabyl()` from `janitor` to produce and “adorn” tabulations and cross-tabulations
- Use `get_summary_stats()` from `rstatix` to easily generate data frames of numeric summary statistics for multiple columns and/or groups
- Use `summarise()` and `count()` from `dplyr` for more complex statistics, tidy data frame outputs, or preparing data for `ggplot()`
- Use `tbl_summary()` from `gtsummary` to produce detailed publication-ready tables
- Use `table()` from base R if you do not have access to the above packages

# Import data
```{r import-data-linelist}
# import the linelist
linelist <- import("data/linelist_cleaned.rds")
```

# Browse data

## skimr package

```{r skim-linelist}
## get information about each variable in a dataset 
skim(linelist)
```

## `rstatix` summary statistics
```{r get-summary-stats-with-rstatix}
linelist %>% 
  get_summary_stats(
    age, wt_kg, ht_cm, ct_blood, temp,  # columns to calculate for
    type = "common")                    # summary stats to return
```

# janitor package

The **janitor** packages offers the `tabyl()` function to produce tabulations and cross-tabulations, which can be “adorned” or modified with helper functions to display percents, proportions, counts, etc.

## Simple tabyl

```{r simple-tabyl-age_cat}
linelist %>% 
  tabyl(age_cat) %>% 
  adorn_pct_formatting(affix_sign = FALSE)
```

## Printing the tabyl

```{r simple-tabyl-age-gender}
linelist %>% 
  tabyl(age_cat, gender) %>% 
  adorn_totals(where = "col") %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_pct_formatting() %>% 
  adorn_ns(position = "front") %>% 
  adorn_title(
    row_name = "年龄组",
    col_name = "性别",
    placement = "combined"
  ) %>% 
  flextable::flextable() %>% 
  flextable::autofit()
```

## Use on other table

```{r janitor-dplyr}
linelist %>% 
  count(hospital) %>% 
  adorn_totals(where = "row")
```

## Statistics

You can apply statistical tests on tabyls^[like `chisq.test()` or `fisher.test()` from the **stats** package]。

```{r statistic-in-tabyl, warning=FALSE}
age_by_outcome <- linelist %>% 
  tabyl(age_cat, outcome, show_na = F)

chisq.test(age_by_outcome)
```

# dplyr package

`summarise() `creates a new, summary data frame. If the data are ungrouped, it will return a one-row dataframe with the specified summary statistics of the entire data frame. If the data are grouped, the new data frame will have one row per group.^[_TIP_: The summarise function works with both UK and US spelling (`summarise()` and `summarize()`).]

## Get counts

```{r count-by-group}
linelist %>% 
  count(age_cat, outcome)
```

```{r count-by-groups-widen-table}
linelist %>% 
  count(age_cat, outcome) %>%
  mutate(percent = scales::percent(n / sum(n))) %>% 
  pivot_wider(names_from = "outcome", values_from = c("n", "percent")) %>% 
  flextable::flextable() %>% 
  flextable::autofit()
```

## Plotting

```{r count-by-group-plotting}
linelist %>% 
  count(age_cat, outcome) %>% 
  ggplot() +
    geom_col(
      mapping = aes(
        x = outcome, 
        fill = age_cat, 
        y = n
      )
    )
```

## Summary statistics and conditional statistics

```{r summary-dplyr-table}
summary_table <- linelist %>%                                        # begin with linelist, save out as new object
  group_by(hospital) %>%                                             # group all calculations by hospital
  summarise(                                                         # only the below summary columns will be returned
    cases       = n(),                                                # number of rows per group
    delay_max   = max(days_onset_hosp, na.rm = T),                    # max delay
    delay_mean  = round(mean(days_onset_hosp, na.rm=T), digits = 1),  # mean delay, rounded
    delay_sd    = round(sd(days_onset_hosp, na.rm = T), digits = 1),  # standard deviation of delays, rounded
    delay_3     = sum(days_onset_hosp >= 3, na.rm = T),               # number of rows with delay of 3 or more days
    pct_delay_3 = scales::percent(delay_3 / cases),                   # convert previously-defined delay column to percent 
    max_temp_fvr = max(temp[fever == "yes"], na.rm = T),
    max_temp_no = max(temp[fever == "no"], na.rm = T)
  )

summary_table
```

## Glueing together
```{r beautify-summary-table}
summary_table %>% 
  mutate(delay = str_glue("{delay_mean} ({delay_sd})")) %>%   # combine and format other values
  select(-c(delay_mean, delay_sd)) %>%                        # remove two old columns 
  adorn_totals(where = "row") %>%                             # add total row
  select(
    "Hospital Name"   = hospital,
    "Cases"           = cases,
    "Max delay"       = delay_max,
    "Mean (sd)"       = delay,
    "Delay 3+ days"   = delay_3,
    "% delay 3+ days" = pct_delay_3    
  ) %>% 
  flextable::flextable() %>% 
  flextable::autofit() # print

```

## Percentiles

```{r simple-percentiles}
# get default percentile values of age (0%, 25%, 50%, 75%, 100%)
linelist %>% 
  summarise(age_percentiles = quantile(age_years, na.rm = TRUE))

# get manually-specified percentile values of age (5%, 50%, 75%, 98%)
linelist %>% 
  summarise(
    age_percentiles = quantile(
      age_years,
      probs = c(.05, 0.5, 0.75, 0.98), 
      na.rm=TRUE)
    )
```

```{r percentile-by-group}
linelist %>% 
  group_by(hospital) %>% 
  summarise(
    p05 = quantile(age_years, probs = 0.05, na.rm=T),
    p50 = quantile(age_years, probs = 0.5, na.rm=T),
    p75 = quantile(age_years, probs = 0.75, na.rm=T),
    p98 = quantile(age_years, probs = 0.98, na.rm=T)
  )
```
all the summary statistics can be produced with get_summary_stat() from the rstatix package.

```{r rstatix-summary-statistics}
linelist %>% 
  group_by(hospital) %>% 
  rstatix::get_summary_stats(age, type = "quantile")
```

## Summarise aggregated data

```{r summarise-aggregated-data-1}
linelist_agg <- linelist %>% 
  drop_na(gender, outcome) %>% 
  count(outcome, gender)

linelist_agg %>% 
  pivot_wider(names_from = "gender", values_from = "n") %>% 
  mutate(
    total_cases = f + m
  ) %>% 
  select(
    outcome, total_cases,
    "male_cases" = m,
    "female_cases"= f
  )
```

```{r summarise-aggregated-data-2}
linelist_agg %>% 
  group_by(outcome) %>% 
  summarise(
    total_cases  = sum(n, na.rm=T),
    male_cases   = sum(n[gender == "m"], na.rm=T),
    female_cases = sum(n[gender == "f"], na.rm=T)
  )
```

## `across()` multiple columns

when you want to calculate the same statistics for many columns

```{r across-multi-col-with-mean-sd}
linelist %>% 
  group_by(outcome) %>% 
  summarise(across(.cols = c(age_years, temp, wt_kg, ht_cm), # columns
                   .fns = list("mean" = mean, "sd" = sd),    # multiple functions 
                   na.rm=T))                                 # extra arguments
```
```{r across-multi-col-with-tidyselect}
linelist %>% 
  group_by(outcome) %>% 
  summarise(across(
    .cols = where(is.numeric),  # all numeric columns in the data frame
    .fns = mean,
    na.rm=T))
```

## Total rows

```{r use-janitor-adorn}
linelist %>% 
  group_by(gender) %>%
  summarise(
    known_outcome = sum(!is.na(outcome)),           # Number of rows in group where outcome is not missing
    n_death  = sum(outcome == "Death", na.rm=T),    # Number of rows in group where outcome is Death
    n_recover = sum(outcome == "Recover", na.rm=T), # Number of rows in group where outcome is Recovered
  ) %>% 
  adorn_totals() %>%                                # Adorn total row (sums of each numeric column)
  adorn_percentages("col") %>%                      # Get column proportions
  adorn_pct_formatting() %>%                        # Convert proportions to percents
  adorn_ns(position = "front")                      # display % and counts (with counts in front)
```

## `summarise()` on “total” data and then `bind_rows()`

```{r bind-total-into-long-table}
by_hospital <- linelist %>% 
  filter(!is.na(outcome) & hospital != "Missing") %>%  # Remove cases with missing outcome or hospital
  group_by(hospital, outcome) %>%                      # Group data
  summarise(                                           # Create new summary columns of indicators of interest
    N = n(),                                           # Number of rows per hospital-outcome group     
    ct_value = median(ct_blood, na.rm=T))              # median CT value per group
  
totals <- linelist %>% 
      filter(!is.na(outcome) & hospital != "Missing") %>%
      group_by(outcome) %>%                            # Grouped only by outcome, not by hospital    
      summarise(
        N = n(),                                       # These statistics are now by outcome only     
        ct_value = median(ct_blood, na.rm=T))

table_long <- bind_rows(by_hospital, totals) %>% 
  mutate(hospital = replace_na(hospital, "Total"))

table_long <- table_long %>% 
  # Pivot wider and format
  ########################
  pivot_wider(                                         # Pivot from long to wide
    values_from = c(ct_value, N),                      # new values are from ct and count columns
    names_from = outcome) %>%                          # new column names are from outcomes
  mutate(                                              # Add new columns
    N_Known = N_Death + N_Recover,                               # number with known outcome
    Pct_Death = scales::percent(N_Death / N_Known, 0.1),         # percent cases who died (to 1 decimal)
    Pct_Recover = scales::percent(N_Recover / N_Known, 0.1)) %>% # percent who recovered (to 1 decimal)
  select(                                              # Re-order columns
    hospital, N_Known,                                 # Intro columns
    N_Recover, Pct_Recover, ct_value_Recover,          # Recovered columns
    N_Death, Pct_Death, ct_value_Death)  %>%           # Death columns
  arrange(N_Known)                                     # Arrange rows from lowest to highest (Total row at bottom)
```

## print table with flextable 

```{r beautify-table-with-flextable}
# define style for border line
border_style = officer::fp_border(color="black", width=1)

table_long %>% 
  flextable() %>% 
  # autofit() %>%
  
  width(j=1, width = 2.7) %>% 
  width(j=2, width = 1.5) %>% 
  width(j=c(4,5,7,8), width = 1) %>% 
  
  add_header_row(
    top = TRUE,                # New header goes on top of existing header row
    values = c("Hospital",     # Header values for each column below
               "Total cases with known outcome", 
               "Recovered",    # This will be the top-level header for this and two next columns
               "",
               "",
               "Died",         # This will be the top-level header for this and two next columns
               "",             # Leave blank, as it will be merged with "Died"
               "")
  ) %>% 
    
  set_header_labels(         # Rename the columns in original header row
    hospital = "", 
    N_Known = "",                  
    N_Recover = "Total",
    Pct_Recover = "% of cases",
    ct_value_Recover = "Median CT values",
    N_Death = "Total",
    Pct_Death = "% of cases",
    ct_value_Death = "Median CT values"
  ) %>% 
  
  merge_at(i = 1, j = 3:5, part = "header") %>%   # Horizontally merge columns 3 to 5 in new header row
  merge_at(i = 1, j = 6:8, part = "header") %>%   # Horizontally merge columns 6 to 8 in new header row
  
  border_remove() %>%    # Remove all existing borders
  
  theme_booktabs() %>%   # add horizontal lines via a pre-determined theme setting
  
  # add vertical lines to separate Recovered and Died sections
  vline(part = "all", j = 2, border = border_style) %>%   # at column 2 
  vline(part = "all", j = 5, border = border_style) %>%   # at column 5  

  # set cell align
  align(align = "center", j = c(2:8), part = "all") %>% 

  # set font  
  fontsize(i = 1, size = 12, part = "header") %>%   # adjust font size of header
  bold(i = 1, bold = TRUE, part = "header") %>%     # adjust bold face of header
  bold(i = 6, bold = TRUE, part = "body") %>%       # adjust bold face of total row (row 7 of body)

  # merge cells
  merge_at(i = 1:2, j = 1, part = "header") %>% 
  merge_at(i = 1:2, j = 2, part = "header") %>% 
  
  # set background color
  bg(part = "body", bg = "gray95") %>% 
  
  # conditional cell color
  bg(., i= ~ hospital == "Military Hospital", part = "body", bg = "#91c293")  %>%
  bg(j = 7, i = ~ Pct_Death >= 55, part = "body", bg = "red") 
```

# *gtsummary* package

use the gtsummary package and its function tbl_summary()

## Summary table

```{r gtsummary-default-setting}
linelist %>% 
  select(age_years, gender, outcome, fever, temp, hospital) %>%  # keep only the columns of interest
  tbl_summary()                                                  # default
```

```{r more-complex-gtsummary}
linelist %>% 
  select(age_years, gender, outcome, fever, temp, hospital) %>% # keep only columns of interest
  tbl_summary(     
    by = outcome,                                               # stratify entire table by outcome
    statistic = list(all_continuous() ~ "{mean} ({sd})",        # stats and format for continuous columns
                     all_categorical() ~ "{n} / {N} ({p}%)"),   # stats and format for categorical columns
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical() ~ "categorical",                # force all categorical levels to display
    label  = list(                                              # display labels for column names
      outcome   ~ "Outcome",                           
      age_years ~ "Age (years)",
      gender    ~ "Gender",
      temp      ~ "Temperature",
      hospital  ~ "Hospital"),
    missing_text = "Missing"                                    # how missing values should display
  )
```

## Multi-line stats for continuous variables 
by setting the $type =$ to “continuous2”.

```{r multi-line-stats-with-gtsummary}
linelist %>% 
  select(age_years, temp) %>%                      # keep only columns of interest
  tbl_summary(                                     # create summary table
    type = all_continuous() ~ "continuous2",       # indicate that you want to print multiple statistics 
    statistic = all_continuous() ~ c(
      "{mean} ({sd})",                             # line 1: mean and SD
      "{median} ({p25}, {p75})",                   # line 2: median and IQR
      "{min}, {max}")                              # line 3: min and max
  )
```

