---
title: "Simple statistical tests"
subtitle: "EPIR Handbook - Chapter 18"
author: "Lin Yong (yong.lin@datarx.cn)"
date: "9/14/2021"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scrool: false
    theme: readable
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(htmltools.dir.version = FALSE)
```

This chapter demonstrates how to conduct simple statistical tests using base R, `rstatix`, and `gtsummary`.

- T-test
- Shapiro-Wilk test
- Wilcoxon rank sum test
- Kruskal-Wallis test
- Chi-squared test
- Correlations between numeric variables

Advantages and disadvantages:

- Use base R functions to print a statistical outputs to the R Console
- Use `rstatix` functions to return results in a data frame, or if you want tests to run by group
- Use `gtsummary` if you want to quickly print publication-ready tables

# Prepare and import data

```{r prepare-library}
pacman::p_load(
  rio,          # File import
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  gtsummary,    # summary statistics and tests
  rstatix,      # summary statistics and statistical tests
  corrr,        # correlation analysis for numeric variables
  janitor,      # adding totals and percents to tables
  scales,       # easily convert proportions to percents  
  flextable     # converting tables to pretty images
  )
```

```{r import-data-linelist}
# import the linelist
linelist <- import("data/linelist_cleaned.rds")
```

# base R

The commands are relatively simple and results will print to the R Console for simple viewing.

## T-tests (Student's t-test)

```{r two-sample-t-test-1}
linelist %>% 
  select(age_years, gender) %>% # keep only columns of interest
  tbl_summary(     
    by = gender,                                               # stratify entire table by outcome
    statistic = all_continuous() ~ "{mean}",        # stats and format for continuous columns    
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical() ~ "categorical",                # force all categorical levels to display
    label  = age_years ~ "Age (years)"                       # display labels for column names
  ) %>% 
  add_p()

## compare mean age by gender group with a t-test
t.test(age_years ~ gender, data = linelist)
```

```{r two-sample-t-test-2}
linelist %>% 
  select(age_years, outcome) %>% # keep only columns of interest
  tbl_summary(     
    by = outcome,                                               # stratify entire table by outcome
    statistic = all_continuous() ~ "{mean}",        # stats and format for continuous columns    
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    label  = age_years ~ "Age (years)",                          # display labels for column names
    missing_text = "Missing"                                    # how missing values should display
  ) %>% 
  add_p()

## compare mean age by outcome group with a t-test
t.test(age_years ~ outcome, data = linelist)
```
```{r one-sample-t-test}
t.test(linelist$age_years, mu = 15, alternative = "greater")
```

The Shapiro-Wilk test can be used to determine whether a sample came from a normally-distributed population (an assumption of many other tests and analysis, such as the t-test). However, this can only be used on a sample between 3 and 5000 observations. For larger samples a quantile-quantile plot may be helpful.

```{r shapior-Wilk test}
shapiro.test(linelist$age_years)
```


# `rstatix` package

Run statistical tests and retrieve results in a “pipe-friendly” framework. The results are automatically in a data frame.

## Summary statistics

```{r get-summary-stats-with-group-data}
linelist %>%
  group_by(hospital) %>%
  rstatix::get_summary_stats(age, temp, type = "five_number")
```

## T-test
```{r}
# Use a formula syntax to specify the numeric and categorical columns
linelist %>% 
  t_test(age_years ~ gender)

linelist %>% 
  t_test(age_years ~ outcome)

# Or use ~ 1 and specify mu = for a one-sample T-test
linelist %>% 
  t_test(age_years ~ 1, mu = 20, alternative = "greater")

# done by group
linelist %>% 
  group_by(gender) %>% 
  t_test(age_years ~ 1, mu = 18, alternative = "greater")
```
## Chi-squared test

```{r Chi-squared test}
linelist %>% 
  tabyl(gender, outcome) %>% 
  select(-1) %>% 
  chisq_test()
```

# `gtsummary` package

## T-tests

```{r}
# Compare the difference in means for a continuous variable in two groups. 

linelist %>% 
  select(age_years, outcome) %>%             # keep variables of interest
  tbl_summary(                               # produce summary table
    statistic = age_years ~ "{mean} ({sd})", # specify what statistics to show
    by = outcome) %>%                        # specify the grouping variable
  add_p(age_years ~ "t.test")                # specify what tests to perform
```


# Correlations

Correlation between numeric variables can be investigated using the tidyverse
corrr package.

```{r correlations}
correlation_tab <- linelist %>% 
  select(generation, age, ct_blood, days_onset_hosp, wt_kg, ht_cm) %>%   # keep numeric variables of interest
  correlate() %>%       # create correlation table (using default pearson)
  shave()

correlation_tab %>% 
  flextable()

# plot correlations
rplot(correlation_tab)
```

