---
title: "Dashboards with R Markdown"
subtitle: "EPIR Handbook - Chapter 27"
author: "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scrool: no
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
pacman::p_load(
  rio,             # data import/export     
  here,            # locate files
  tidyverse,       # data management and visualization
  flexdashboard,   # dashboard versions of R Markdown reports
  shiny,           # interactive figures
  plotly           # interactive figures
)

knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(htmltools.dir.version = FALSE)
```

# Import data

Import the dataset of cases from a simulated Ebola epidemic:

```{r cars}
# import the linelist
linelist <- import("data/linelist_cleaned.rds")

head(linelist)
```

# Create a new R Markdown dashboard

Select “From Template” and select the “Flex Dashboard” template

```{r eval=FALSE}
rmarkdown::draft("outbreak_dashboard.Rmd",
                 template = "flex_dashboard",
                 package = "flexdashboard")
```

## Headings

In flexdashboard, a primary heading (#) creates a “page” of the dashboard. Second-level headings (##) create a column or a row depending on your orientation: parameter (see details below). Third-level headings (###) create panels for plots, charts, tables, text, etc.

## Adding content




