---
title: "Ebola Outbreak Dashboard"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
runtime: shiny
---

```{r global, include=FALSE}
pacman::p_load(
  rio,             # data import/export     
  here,            # locate files
  tidyverse,       # data management and visualization
  flexdashboard,   # dashboard versions of R Markdown reports
  flextable,
  incidence2,
  epicontacts,
  DT,
  janitor,
  shiny,           # interactive figures
  plotly,          # interactive figures
  tsibble
)

# import the linelist
linelist <- import(here("data", "linelist_cleaned.rds"))
```

Outbreak dashboard {.sidebar}
-----------------------------------------------------------------------

This is a sidebar that can be viewed from all pages.

You could just have text here.

Or if using **shiny** with **flexdashboard**, you might put your input controls here.


Column 1 {data-width=350}
-----------------------------------------------------------------------

### Summary and actions items
 This report is for the Incident COmmand team of the fictional outbreak of Ebola cases. **As of `r format(max(linelist$date_hospitalisation, na.rm=T), "%d %B")` there have been `r nrow(linelist)` cases reported as hospitalized.**
 
 * Several previously-unaffected areas to the West are now reporting cases
 * Internal reviews suggest that better communication is needed between district and provincial level data management teams
 * Safe and dignified burial teams are reporting difficulties


### Review data

#### Cases by hospitals

```{r}
linelist %>% 
  count(hospital) %>% 
  adorn_totals() %>% 
  rename("Hospital" = hospital, 
         "Cases" = n) %>%
  knitr::kable()
```

#### Display Epicurve of cases admitted to:

```{r}
hospital_choices <- rev(unique(linelist$hospital))

selectInput("hospital_choice", label = "",
            choices = hospital_choices, 
            selected = hospital_choices[[1]])
```

Column 2 
-----------------------------------------------------------------------

### Epdimic curve by age

```{r}
renderPlot({
  age_outbreak <- incidence(
    linelist %>% filter(hospital == input$hospital_choice),
    date_index = date_onset, 
    interval = "week", 
    groups = age_cat)
  
  plot(age_outbreak, fill = age_cat, col_pal = muted, 
       title = str_glue("Cases admitted to {input$hospital_choice}"))
})
```

