---
title: "R4DS Vectors"
subtitle: "EPIR Handbook - Chapter 17"
author: 
  - "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scrool: false
    theme: paper
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(
  tidyverse    # data management + ggplot2 graphics 
  )
```

## Vector basics

1. Its **type**
```{r}
typeof(letters)
typeof(1:10)
```

2. Its **lenth**
```{r}
x <- list("a", "b", 1:10)
length(x)
```

**augmented vectors**
- Factors (on top of integer vectors)
- Dates and date-times (on top of numeric vectors)
- Data frames and tibbles (on top of lists)

## Type of atomic vector

### Logical
Three possible values: FALSE, TRUE, NA.

### Numeric
Numeric vector: Integer and double.

Numbers are double by default. To make an integer, place an `L` after the number.

```{r}
typeof(1)
typeof(1L)
```
Tow important differences:

1. Doubles are approximations.
2. Integers have `NA`. Doubles have: `NA`, `NaA`, `Inf` and `-Inf`

```{r}
c(-1, 0, 1) / 0
```
Checking for those special values: `is.finite()`, `is.infinite()`, `is.na()` and `is.nan()`.

### Character
R uses a global string pool. This means that each unique string is only stored in memory once, and every use of the string points to that representation. 


## Using atomic vectors

- Explicit coercion
- Implicit coercion











