---
title: "Chapter 3 Bar Graphs"
subtitle: "R Graphics Cookbook - Chapter 3"
author: "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scrool: no
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

pacman::p_load(
  gcookbook,    # Load gcookbook for the pg_mean data set
  tidyverse
)
```

## Making a Basic Bar Graph

```{r}
ggplot(pg_mean, aes(x = group, y = weight)) +
  geom_col()
```

By default, bar graphs use a dark grey for the bars. To use a color fill, use `fill`. Also, by default, there is no outline around the fill. To add an outline, use `colour`. 

```{r}
ggplot(pg_mean, aes(x = group, y = weight)) +
  geom_col(fill = "lightblue", color = "black")
```

## Grouping Bars Together

Use the cabbage_exp data set, which has two categorical variables, `Cultivar` and `Date`, and one continuous variable, `Weight`:


```{r}
knitr::kable(head(cabbage_exp))

cabbage_exp %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(position = "dodge", colour = "black") +
    scale_fill_brewer(palette = "Pastel1")
```

## Making a Bar Graph of Counts

Use `geom_bar()` without mapping anything to `y`:

```{r}
# Equivalent to using geom_bar(stat = "bin")
diamonds %>% 
  ggplot(aes(x = cut)) +
    geom_bar()
```

```{r}
ggplot(diamonds, aes(x = carat, fill = cut)) +
  geom_histogram()
```


## Using Colors in a Bar Graph

Map the appropriate variable to the fill aesthetic. Note that setting occurs outside of `aes()`, while mapping occurs within `aes()`:

```{r}
upc <- uspopchange %>%
  arrange(desc(Change)) %>%
  slice(1:15)

upc %>% 
  ggplot(aes(x = reorder(Abb, Change), 
             y = Change, 
             fill = Region)) +
    geom_col(colour = "black") +
    scale_fill_manual(values = c("#669933", "#FFCC66")) +
    xlab("State")
```

## Coloring Negative and Positive Bars Differently

```{r}
climate_sub <- climate %>%
  filter(Source == "Berkeley" & Year >= 1900) %>%
  mutate(pos = Anomaly10y >= 0)
```

```{r}
climate_sub %>% 
  ggplot(aes(x = Year, y = Anomaly10y, fill = pos)) +
    geom_col(position = "identity", colour = "black", size = 0.25) +
    scale_fill_manual(values = c("#CCEEFF", "#FFDDDD"), guide = "none")
```


## Adjusting Bar Width and Spacing

To make the bars narrower or wider, set width in geom_col(). The default value is 0.9; larger values make the bars wider, and smaller values make the bars narrower:

```{r}
pg_mean %>% 
  ggplot(aes(x = group, y = weight)) +
    geom_col(width = 0.5)
```

For grouped bars, the default is to have no space between bars within each group. To add space between bars within a group, make width smaller and set the value for `position_dodge` to be larger than width:

```{r}
cabbage_exp %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(width = 0.6, position = position_dodge(0.7))
```


## Making a Stacked Bar Graph

Use `geom_bar()` and map a variable `fill`.

```{r}
cabbage_exp %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(width = 0.6, colour = "black") +
    scale_fill_brewer(palette = "Pastel1")
```

Reverse the legend and stack:

```{r}
cabbage_exp %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(width = 0.6, 
             position = position_stack(reverse = TRUE)) +
    guides(fill = guide_legend(reverse = TRUE))
```

## Making a Proportional Stacked Bar Graph

Make a stacked bar graph that shows proportions (also called a 100% stacked bar graph). Use `geom_col(position = "fill")`. With `position = "fill"`, the `y` values will be scaled to go from 0 to 1. To print the labels as percentages, use `scale_y_continuous(labels = scales::percent)`

```{r}
cabbage_exp %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(width = 0.7, 
             colour = "black", 
             position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Pastel1")
```

## Adding Labels to a Bar Graph

Add `geom_text()` to your graph. It requires a mapping for `x`, `y`, and the text itself. By setting `vjust` (the vertical justification), it is possible to move the text above or below the tops of the bars:

```{r}
cabbage_exp %>% 
  ggplot(aes(x = interaction(Date, Cultivar), y = Weight)) +
    geom_col() +
    geom_text(aes(label = Weight), vjust = 1.5, colour = "white")
```

Putting labels on stacked bar graphs requires finding the cumulative sum for each stack. To do this, first make sure the data is sorted properly – if it isn’t, the cumulative sum might be calculated in the wrong order. We’ll use the `arrange()` function from the `dplyr` package. 

```{r}
# Sort by the Date and Cultivar columns
ce <- cabbage_exp %>%
  arrange(Date, rev(Cultivar)) %>% 
  group_by(Date) %>%
  mutate(label_y = cumsum(Weight) - 0.5 * Weight)

ce %>% 
  ggplot(aes(x = Date, y = Weight, fill = Cultivar)) +
    geom_col(colour = 'black') +
    geom_text(aes(y = label_y, 
                  label = paste(format(Weight, nsmall = 2), 
                                "kg")), 
              size = 4) +
  scale_fill_brewer(palette = "Pastel1")
```

## Making a Cleveland Dot Plot

Cleveland dot plots are an alternative to bar graphs that reduce visual clutter and can be easier to read.

The simplest way to create a dot plot is to use `geom_point()`:

```{r}
# Take the top 25 from the tophitters data set
tophit <- tophitters2001[1:25, ] 

ggplot(tophit, aes(x = avg, y = name)) +
  geom_point()
```

By default, the items on the given axis will be ordered however is appropriate for the data type. A character vector is ordered alphabetically. If it were a factor, it would use the order defined in the factor levels. 

```{r}
ggplot(tophit, aes(x = avg, y = reorder(name, avg))) +
  geom_point(size = 3) +  # Use a larger dot
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey60", linetype = "dashed")
  )
```

```{r}
ggplot(tophit, aes(x = reorder(name, avg), y = avg)) +
  geom_point(size = 3) +  # Use a larger dot
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey60", linetype = "dashed"),
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

```{r}
# Get the names, sorted first by lg, then by avg
nameorder <- tophit$name[order(tophit$lg, tophit$avg)]

# Turn name into a factor, with levels in the order of nameorder
tophit$name <- factor(tophit$name, levels = nameorder)

ggplot(tophit, aes(x = avg, y = name)) +
  geom_segment(aes(yend = name), xend = 0, colour = "grey50") +
  geom_point(size = 3, aes(colour = lg)) +
  scale_colour_brewer(palette = "Set1", limits = c("NL", "AL")) +
  theme_bw() +
  theme(
    # No horizontal grid lines
    panel.grid.major.y = element_blank(),   
    legend.position = c(1, 0.55),
    # Put legend inside plot area
    legend.justification = c(1, 0.5)
  )
```

Another way to separate the two groups is to use facets.

```{r}
tophit %>%  
  ggplot(aes(x = avg, y = name)) +
    geom_segment(aes(yend = name), xend = 0, 
                 colour = "grey50") +
    geom_point(size = 3, aes(colour = lg)) +
    scale_colour_brewer(palette = "Set1", 
                        limits = c("NL", "AL"), 
                        guide = FALSE) +
    theme_bw() +
    theme(panel.grid.major.y = element_blank()) +
    facet_grid(lg ~ ., scales = "free_y", space = "free_y")
```

