---
title: "Chapter 4 Line Graph"
subtitle: "R Graphics Cookbook - Chapter 4"
author: "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scrool: no
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

pacman::p_load(
  gcookbook,    # Load gcookbook for the pg_mean data set
  tidyverse
)
```

## Basic Line Graph
```{r}
BOD %>% ggplot(aes(x = Time, y = demand)) +
  geom_line()
```
Line graphs can be made with discrete (categorical) or continuous (numeric) variables on the x-axis. When the x variable is a factor, you must also use `aes(group=1)` to ensure that ggplot knows that the data points belong together and should be connected with a line.

With ggplot2, the default y range of a line graph is just enough to include the y values in the data. For some kinds of data, it’s better to have the y range start from zero. You can use `ylim()` to set the range, or you can use `expand_limits()` to expand the range to include a value. 

```{r}
# These have the same result
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line() +
  ylim(0, max(BOD$demand))

# ggplot(BOD, aes(x = Time, y = demand)) +
#   geom_line() +
#   expand_limits(y = 0)
```

## Adding Points to a Line Graph

Add `geom_point()`:

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line() +
  geom_point()
```

Sometimes it is useful to indicate each data point on a line graph. This is helpful when the density of observations is low, or when the observations do not happen at regular intervals. 

```{r}
ggplot(worldpop, aes(x = Year, y = Population)) +
  geom_line() +
  geom_point()

# Same with a log y-axis
ggplot(worldpop, aes(x = Year, y = Population)) +
  geom_line() +
  geom_point() +
  scale_y_log10()
```

## Making a Line Graph with Multiple Lines

In addition to the variables mapped to the x- and y-axes, map another (discrete) variable to colour or linetype.

```{r}
# Map supp to colour
ggplot(tg, aes(x = dose, y = length, colour = supp)) +
  geom_line()

# Map supp to linetype
ggplot(tg, aes(x = dose, y = length, linetype = supp)) +
  geom_line()
```

> Note
>
> If the x variable is a factor, you must also tell ggplot to group by that same variable, as described below.

Line graphs can be used with a continuous or categorical variable on the x-axis. Sometimes the variable mapped to the x-axis is conceived of as being categorical, even when it’s stored as a number. 

```{r}
ggplot(tg, aes(x = factor(dose), y = length, 
               colour = supp, group = supp)) +
  geom_line()
```

If your plot has points along with the lines, you can also map variables to properties of the points, such as shape and fill:

```{r}
# ggplot(tg, aes(x = dose, y = length, shape = supp)) +
  # geom_line() +
  # geom_point(size = 4)  # Make the points a little larger

ggplot(tg, aes(x = dose, y = length, fill = supp)) +
  geom_line() +
  geom_point(size = 4, shape = 21)  # Also use a point with a color fill
```

## Changing the Appearance of Lines

The type of line (solid, dashed, dotted, etc.) is set with `linetype`, the thickness (in mm) with `size`, and the color of the line with `colour` (or `color`).

These properties can be set by passing them values in the call to geom_line():

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line(linetype = "dashed", size = 1, colour = "blue")
```

The default colors aren’t the most appealing, so you may want to use a different palette:

```{r}
ggplot(tg, aes(x = dose, y = length, colour = supp)) +
  geom_line() +
  scale_colour_brewer(palette = "Set1")
```

## Changing the Appearance of Points

In `geom_point()`, set the `size`, `shape`, `colour`, and/or `fill` outside of `aes()`:

```{r}
ggplot(BOD, aes(x = Time, y = demand)) +
  geom_line() +
  geom_point(size = 4, shape = 22, colour = "darkred", fill = "pink")
```

The default shape for points is a solid circle, the default size is 2, and the default colour is black. The fill color is relevant only for some point shapes (numbered 21–25), which have separate outline and fill colors. The fill color is typically `NA`, or empty; you can fill it with white to get hollow-looking circles.

If the points and lines have different colors, you should specify the points after the lines, so that they are drawn on top. Otherwise, the lines will be drawn on top of the points.

The default colors are not very appealing, so you may want to use a different palette, using `scale_colour_brewer()` or `scale_colour_manual()`. To set a single constant shape or size for all the points, specify shape or size outside of `aes()`:

```{r}
# Save the position_dodge specification because we'll use it multiple times
pd <- position_dodge(0.2)

ggplot(tg, aes(x = dose, y = length, fill = supp)) +
  geom_line(position = pd) +
  geom_point(shape = 21, size = 3, position = pd) +
  scale_fill_manual(values = c("black","white"))
```

## Making a Graph with a Shaded Area

Use `geom_area()` to get a shaded area:

```{r}
# Convert the sunspot.year data set into a data frame for this example
sunspotyear <- data.frame(
    Year     = as.numeric(time(sunspot.year)),
    Sunspots = as.numeric(sunspot.year)
  )

sunspotyear %>% 
  ggplot(aes(x = Year, y = Sunspots)) +
    geom_area()
```

By default, the area will be filled with a very dark grey and will have no outline. The color can be changed by setting fill. In the following example, we’ll set it to "blue", and we’ll also make it 80% transparent by setting alpha to 0.2. This makes it possible to see the grid lines through the area. We’ll also add an outline, by setting colour:

```{r}
ggplot(sunspotyear, aes(x = Year, y = Sunspots)) +
  geom_area(colour = "black", fill = "blue", alpha = .2)
```

## Making a Stacked Area Graph

Use geom_area() and map a factor to fill:

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area()
```

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(colour = "black", size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues")
```

## Making a Proportional Stacked Area Graph

Use `geom_area(position = "fill")`:

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(position = "fill", colour = "black", 
            size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues")
```

With position="fill", the y values will be scaled to go from 0 to 1. To print the labels as percentages, use scale_y_continuous(labels = scales::percent):

```{r}
ggplot(uspopage, aes(x = Year, y = Thousands, fill = AgeGroup)) +
  geom_area(position = "fill", colour = "black", 
            size = .2, alpha = .4) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(labels = scales::percent)
```

## Adding a Confidence Region

Use `geom_ribbon()` and map values to `ymin` and `ymax`.

```{r}
# Grab a subset of the climate data
climate_mod <- climate %>%
  filter(Source == "Berkeley") %>%
  select(Year, Anomaly10y, Unc10y)

# Shaded region
climate_mod %>% 
  ggplot(aes(x = Year, y = Anomaly10y)) +
    geom_ribbon(aes(ymin = Anomaly10y - Unc10y, ymax = Anomaly10y + Unc10y), alpha = 0.2) +
    geom_line()
```

```{r}
ggplot(climate, aes(x = Year, y = Anomaly10y, colour = Source)) +
  geom_line() +
  scale_colour_brewer(palette = "Set1")
```
