---
title: "Chapter 17 - Comparing Two Measures of Centrality"
subtitle: "RMRWR - Reproducible Medical Research with R"
author: 
  - "Peter D.R. Higgins, MD, PhD, MSc"
  - "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scrool: false
    theme: readable
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
pacman::p_load(
  rio,          # File import
  here,         # File locator
  skimr,        # get overview of data
  tidyverse,    # data management + ggplot2 graphics 
  gtsummary,    # summary statistics and tests
  rstatix,      # summary statistics and statistical tests
  janitor,      # adding totals and percents to tables
  scales,       # easily convert proportions to percents  
  flextable,    # converting tables to pretty images
  medicaldata,  # medical data
  car           # qqplot
  )

knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(htmltools.dir.version = FALSE)
```

Comparing two groups
- Mean or median vs. expected
- Two arms of study - independent
- Pre and post / spouse and partner / left vs right arm – paired groups
- Are the means significantly different?
- Or the medians (if not normally distributed)?

## The data

```{r importing}
data <- cytomegalovirus

# skim the data set
skim(data)

data %>% 
  group_by(sex) %>% 
  summarize(mean_time2cmv = mean(time.to.cmv))
```

## Visulization

```{r}
data %>% 
  ggplot(mapping = aes(time.to.cmv)) +
  geom_density() +
  # facet_wrap(~sex) +
  theme_linedraw()
```
```{r}
data %>% 
  get_summary_stats(
    time.to.cmv, CD3.dose,  # columns to calculate for
    type = "common")                    # summary stats to return

```

```{r message=FALSE, warning=FALSE}
data %>% 
  ggplot(mapping = aes(CD3.dose)) +
  # geom_histogram() +
  geom_density() +
  facet_wrap(~sex) +
  theme_linedraw()
```

```{r normality-test}
data$time.to.cmv %>% 
  shapiro.test()

car::qqPlot(data$time.to.cmv, ylab="Time to CMV", main="Time to CMV qq-plot")
```

## Mammal sleep hours {.tabset}

```{r mammal-sleep-data}
df <- msleep

skim(df)
```

### Density

```{r sleeptotal-density}
df %>% 
  ggplot(mapping = aes(df$sleep_total)) +
    geom_density() +
    theme_linedraw()
```

### Shapiro Test

```{r sleeptotal-normality-test}
shapiro.test(df$sleep_total)
#df %>% shapiro_test(sleep_total)
```

### qqPlot

```{r sleeptotal-qqplot}
car::qqPlot(df$sleep_total, ylab="Total Sleep time", main="Total Sleep qq-plot")
```

## One Sample T test

Is the mean roughly 8 hours of sleep per day?

univariate test

- $H_0$: mean is 8 hours
- $H_1$: mean is not 8 hours

```{r slepptotal-t-test-8h}
t.test(df$sleep_total, mu = 8)
```

## 2 group t test  {.tabset}

- hypothesis - herbivores need more time to get food, sleep less than carnivores
- how to test this?
  - normal, so can use t test for 2 groups

```{r}
df %>% 
  filter(vore %in% c("herbi", "carni")) %>% 
  select(name, vore, sleep_total) %>% 
  ggplot(mapping = aes(sleep_total)) +
    geom_density() +
    facet_wrap(~vore) +
    theme_linedraw()  
```
  
### Results

```{r}
result <- df %>% 
  filter(vore %in% c("herbi", "carni")) %>% 
  t.test(formula = sleep_total ~ vore, data = .)
  
result

result %>% 
  tidy() %>% 
  flextable()
```

### Interpreting

- Welch t-test (not Student)
  - Welch does NOT assume equal variances in each group
- p value NS
- accept null hypothesis
  - Ho: means of groups roughly equal
  - Ha: means are different
  - 95% CI crosses 0
- Carnivores sleep a little more, but not a lot
