---
title: "03 - Tasting Menu of R"
subtitle: "RMRWR - Reproducible Medical Research with R"
author: 
  - "Peter D.R. Higgins, MD, PhD, MSc"
  - "Lin Yong (yong.lin@datarx.cn)"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scrool: false
    theme: readable
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(rstatix)
library(medicaldata)
library(glue)
prostate <- medicaldata::blood_storage %>% clean_names()

theme <- theme_get()
#using the Chinese fonts you have, check it with font book.  
theme$text$family <- "STSong"
theme_set(theme)
```

## Analysis of Prostate Data

Wrangling the Prostate data.  We will be modifying the prostate dataset (with the mutate function), particularly the variables aa and fam_hx into properly labeled categorical variables, called factors in R. Then we will save (assign) the result to the prostate_factors object, while retaining the previous version in the prostate object.

```{r wrangle}
prostate_factors <- prostate %>%
  mutate(aa = factor(aa, levels = c(0, 1),
                     labels = c("白人", "非裔美国人"))) %>%
  mutate(fam_hx = factor(fam_hx, levels = c(0, 1),
                         labels = c("No Family History", "FHx of Prostate Cancer")))
```

We will run some code to summarize three variables:

```{r summarize, echo=FALSE, message=FALSE, warning=FALSE}
prostate_factors %>% 
  select(age, p_vol, preop_psa, aa, fam_hx) %>% 
  group_by(aa, fam_hx) %>% 
  summarize(across(age:preop_psa, mean, na.rm = TRUE))
```


## Including Plots

```{r visualize, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(prostate_factors) +
  aes(x = p_vol, y = preop_psa, col = aa) +
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_grid(aa ~ fam_hx) +
  labs(x = "Prostate Volume", y = "Preoperative PSA",
       title = "Relationship Between Prostate Volume and Preop PSA, \nSubdivided by Family History and Race") +
  theme(legend.position = "bottom")
```

## Statistical Testing
```{r t-test}
prostate_factors %>% 
  t_test(formula = preop_psa ~ aa,
         detailed = TRUE)
```

## Importing data

```{r scurvy, eval=FALSE, include=FALSE}
url_stem <- "https://raw.githubusercontent.com/higgi13425/rmrwr-book/master/"

scurvy_data <- read_csv(glue(url_stem, 'data/scurvy.csv'))

# scurvy_cols <- read_csv(file = glue(url_stem, 'data/scurvy.csv'),
#                         col_types = cols(
#                           study_id = col_integer(),
#                           treatment = col_factor(),
#                           dosing_regimen_for_scurvy = col_character(),
#                           gum_rot_d6 = col_character(),
#                           skin_sores_d6 = col_character(),
#                           weakness_of_the_knees_d6 = col_character(),
#                           lassitude_d6 = col_character(),
#                           fit_for_duty_d6 = col_character()
#                         ))

scurvy_cols2 <- read_csv(glue(url_stem, 'data/scurvy.csv'),
                         col_types = "ifcffff")
```

```{r strep, eval=FALSE, include=FALSE}
strep_data <- read_tsv(glue(url_stem, 'data/strep_tb.tsv'),
                       col_types = "cfnnfffffccil")

```

## Explore with skimr and DataExplorer

```{r skimr}
library(skimr)

skimr::skim(strep_tb)
```

```{r message=FALSE, warning=FALSE}
# DataExplorer::create_report(strep_tb)

```




